\documentclass[12pt]{article}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs,multirow}
\usepackage{natbib}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}
\usepackage{caption}
\input{tables/numbers.tex}

\title{Forecasting Congressional Votes in the AI Era:\\
An Open Benchmark and an Audit of What Prediction Really Shows}
\author{Andrew B.\ Hall\thanks{Stanford University. Replication materials,
including the full pipeline, model registry, experiment ledger, and
session logs of the AI-assisted development process, are available in the
project repository. All statistics in this paper are generated
programmatically from the analysis code's output files; no number is
transcribed by hand.}}
\date{Draft: July 2026 --- preliminary, comments welcome}

\begin{document}
\maketitle
\onehalfspacing

\begin{abstract}
\noindent A decade ago, a research program at the intersection of machine
learning and legislative studies asked whether the text of a bill could
predict how members of Congress would vote on it
\citep{gerrish2011predicting, gerrish2012they, kraft2016embedding}. We
revisit that program with modern tools and, just as importantly, with
modern evaluation discipline. Using an open benchmark built on 10.6
million individual voting decisions (101st--119th Congresses), we compare
the 2011--2016 architectures to a contemporary stack---sentence-embedding
representations of rollcall-level text, deep parameter heads, member
history offsets, and post-hoc calibration---under four evaluation regimes
of increasing difficulty: random vote completion, random rollcall
holdout, within-congress temporal forecasting, and whole-congress-out
transfer, plus a cryptographically frozen prospective test on votes that
had not occurred when the model was finalized. Three findings emerge.
First, the modern stack forecasts far better than anything prior: test
log loss of \champForecastLL{} versus \gbTcalForecastLL{} for a
calibrated Gerrish--Blei-style model, with \champForecastAcc\% accuracy
and \champContestedAcc\% on contested votes. Second, the earlier
literature's apparent success depended on evaluation choices: under
proper scoring rules and honest splits, no 2011--2016 architecture beats
a two-way count table of member voting rates, and the uncalibrated
classical model is worse than predicting the base rate. Third, the hard
limits are structural, not representational: amendment votes resist
text-based improvement because coalitions form around deals rather than
content, and no model we test carries member behavior across a change in
chamber majority. We release the benchmark, all code, and a
pre-registered experiment ledger documenting negative results alongside
positive ones.
\end{abstract}

\section{Introduction}

How predictable is a member of Congress? The question is old, but it has
acquired new stakes. Measures of legislator behavior---ideal points,
loyalty scores, issue positions---anchor an enormous empirical
literature, and nearly all of them are byproducts of a prediction
problem: a model is fit to rollcall votes, and its parameters are read as
measurements of the members who cast them \citep{poole1997congress,
clinton2004statistical}. The better a model predicts, the more behavioral
structure it has captured, and the more there is to measure. Prediction
quality is thus not an engineering vanity metric; it is an upper bound on
what vote-based measurement can know.

Between 2011 and 2016, a series of papers imported the tools of that
era's machine learning---topic models and early word
embeddings---to push prediction beyond the spatial model, asking whether
a bill's \emph{text} could forecast votes on legislation the model had
never seen \citep{gerrish2011predicting, gerrish2012they,
kraft2016embedding, lauderdale2014scaling}. The program then largely went
quiet, well before the tools that would seem most suited to it existed.
Modern language models read legislative text incomparably better than
LDA. What would the program find today?

This paper answers that question, but with an amendment to the question
itself. The 2011--2016 literature evaluated prediction almost exclusively
on randomly held-out votes, reported accuracy rather than probabilistic
scores, and never tested calibration. We show that these choices were not
innocuous: they flattered the methods of that era and obscured both how
little bill text then contributed and how badly the resulting
probabilities were miscalibrated. Any modern revival has to fix the
yardstick before racing on it. We therefore make the evaluation regime a
first-class object of study. Our benchmark scores every model, classical
and modern, under four regimes of increasing difficulty---random vote
completion, random rollcall holdout, within-congress temporal
forecasting, and whole-congress-out transfer---using proper scoring rules
throughout, and adds a fifth test that no amount of researcher iteration
can contaminate: a frozen, hash-pinned model scored only on votes that
occurred after it was finalized.

Three sets of results follow. \emph{The modern stack is much better.}
Our best model---a blend of towers that read rollcall-level text with
modern sentence encoders, project it through deep parameter heads, and
anchor on member voting histories with temporally calibrated
probabilities---reaches a forecast-regime test log loss of
\champForecastLL{} (accuracy \champForecastAcc\%, contested-vote accuracy
\champContestedAcc\%, AUC \champForecastAUC), against
\gbTcalForecastLL{} for the best-case classical architecture and
\memberqForecastLL{} for the strongest count baseline. The gains
decompose informatively: rollcall-level text construction and the
member-history offset contribute most; swapping a larger text encoder
into the classical architecture buys little; post-hoc temporal
calibration is worth more than a decade of architecture change.

\emph{The old yardstick was broken.} Under proper scoring, the
Gerrish--Blei-style spatial model with text-predicted bill parameters is
\emph{worse than predicting the base rate} in forecast regimes
(log loss \gbRawForecastLL{} against \constantForecastLL{} for the
constant model) despite respectable accuracy---a pure calibration
failure invisible to the metrics of the original papers. Even after we
repair its calibration, it does not beat a member-by-question-type count
table. And contrary to our own expectation, the literature's
random-rollcall holdout was \emph{not} materially easier than a temporal
split once test composition is held fixed: the regime choices that
matter enormously are vote completion versus rollcall holdout, and
within-congress versus cross-congress.

\emph{The remaining failures are structural.} A third of residual loss
sits in amendment votes, and we show---through a sequence of
pre-registered experiments culminating in a placebo-controlled
attribution---that giving models the amendments' actual text does not
help: within a bill, models exploit the \emph{pooling} of an amendment
series more effectively than they can exploit purpose language, because
amendment coalitions form around deals, not content. And when the
majority party changes, everything a model has learned from member
histories partially inverts: pooled member records \emph{anti-transfer}
across a majority flip, our best model's probabilities on an unseen
congress are worse-calibrated than a three-way count table that merely
knows who holds the chamber, and---strikingly---a simple logit on frozen
DW-NOMINATE scores nearly matches the deep model's cross-congress
accuracy (\nomctxCongressoutAcc\% versus \champCongressoutAcc\%).
Prediction inside a stable political environment and prediction across
environments are different problems, and only the first is anywhere near
solved.

Finally, we practice the evaluation discipline we preach. Every
experiment in the development of our champion model was pre-registered
in a public ledger with its hypothesis and falsification test written
before the run; selection used validation data only; the final
comparison observed the test sets exactly once. The headline forecasting
claim rests on the frozen prospective protocol: a model pinned by
checksum in June 2026 scored \prospectiveVoneLL{} log loss
(\prospectiveVoneAcc\% accuracy) on the \prospectiveN{} votes cast in
the three weeks after it was frozen---indistinguishable from its
development test performance, which is precisely what the protocol
exists to demonstrate.

\section{The Prediction Problem and Its Regimes}

The unit of prediction is a (member, rollcall) pair: given everything
observable before the vote, output $P(\text{yea})$. What ``observable''
means is exactly what distinguishes evaluation regimes, and we argue the
literature has been imprecise about it in ways that matter.

\textbf{Completion.} Hold out a random ten percent of individual votes;
other members' votes on the same rollcall remain visible. This is the
psychometric regime in which ideal-point models are traditionally
assessed. It measures representation quality, not forecasting: the
rollcall's own parameters are estimated from its observed votes.

\textbf{Random rollcall holdout.} Hold out entire rollcalls at random;
predict every vote on them from bill features and member history. This
is the regime of \citet{gerrish2011predicting} and
\citet{kraft2016embedding}. Held-out rollcalls are interleaved in time
with training rollcalls.

\textbf{Temporal forecast.} Hold out the final ten percent of rollcalls
within each congress-chamber by date. Nothing contemporaneous with or
subsequent to the evaluation votes is observable. This is the regime a
real forecaster inhabits, up to the stability of the congress itself.

\textbf{Congress-out transfer.} Train through the 116th Congress;
predict the 118th. Member composition, agenda, and---critically---House
majority control all change.

\textbf{Prospective.} Freeze a model, record its hash, and score it only
on votes that occur afterward. Development test sets are observed
repeatedly across a research program and silently absorb researcher
degrees of freedom \citep{dwork2015holdout}; the future cannot.

Throughout, our primary metric is log loss---a proper scoring rule that
rewards calibrated probabilities---with accuracy, AUC, contested-vote
accuracy (rollcalls with minority share of at least 35 percent), and the
NOMINATE literature's APRE reported for comparability. The case for
proper scoring is not pedantry. A downstream user of vote
predictions---a researcher weighting observations, a whip counting
uncertain votes---consumes probabilities, and Section~\ref{sec:audit}
shows that accuracy rankings and probability-quality rankings of the
classical models are not merely different but sometimes inverted.

\section{Data and Benchmark}

We build from Voteview's individual-level records
\citep{lewis2026voteview}: every yea/nay cast in the 101st through 119th
Congresses, 10.6 million decisions after standard exclusions
(present/abstain codes excluded from the target; presidents excluded).
Bill-side features come from GovInfo BILLSTATUS records for the 108th
Congress onward: titles, policy areas, legislative subjects, sponsors,
dated summary versions, and---parsed for this paper---amendment purpose
text linked to specific rollcalls through recorded-vote references
(with a session-numbering correction documented in the appendix).
Rollcall-level text is assembled strictly from information available
before each vote: the vote question, the vote description, and the
latest bill summary version dated before the vote date. Splits are
deterministic, seeded, and content-hashed; the harness strips labels
before models see evaluation rows; count baselines, spatial models,
literature reimplementations, and modern towers all implement one
interface and are scored by one runner. The benchmark, registry, and
every result file in this paper are public.

\section{Models}
\label{sec:models}

\textbf{Count baselines.} Smoothed empirical rates keyed on increasingly
specific cells: the global rate; member rate; party-by-rollcall
majority (completion regime only); party- and member-by-question-type
rates; and, for the transfer regime, rates keyed on \emph{majority
status} rather than congress, which transfer to an unseen congress by
construction.

\textbf{Spatial models.} Logistic ideal points fit by stochastic
gradient descent in one to eight dimensions, and the classical reference
in which member positions are frozen at DW-NOMINATE scores and only
rollcall parameters are estimated.

\textbf{Literature reimplementations.} A Gerrish--Blei-style two-stage
model: ideal points fit on training votes; ridge regression mapping
rollcall text to each rollcall's discrimination and difficulty; unseen
rollcalls scored with text-predicted parameters. We run it with
era-appropriate text features (TF--IDF with latent semantic analysis, the
closest classical analog to their topics) and, separately, with modern
sentence embeddings---isolating what representation alone buys the
classical architecture---and each variant both raw and with temperature
calibration, the latter a courtesy the original inference would have
partially supplied through its priors. A Kraft-style bilinear model
(learned member embeddings against fixed text embeddings) and a logit on
DW-NOMINATE scores interacted with question type and sponsor-party
alignment complete the panel.

\textbf{The modern stack.} Two-tower models in which a member embedding
meets rollcall parameters amortized from text: sentence-embedding
representations of rollcall-level text (question, description, pre-vote
bill summary), a two-layer head mapping text to discrimination and
difficulty, a member-by-question-type empirical-rate offset, and scalar
temperature-and-bias calibration fit on a \emph{temporal} internal slice
of training data---the last ten weeks of the training window, never
validation or test. The final champion is a logistic blend of three such
towers (MiniLM embeddings with a deep head; embeddings plus TF--IDF; a
Qwen-encoder variant), with blend weights fit on the same internal
slice.

Two design lessons deserve emphasis because both were forced on us by
failed alternatives, documented in the experiment ledger. First,
forecast models must early-stop and calibrate on \emph{future-like}
internal data; random in-sample slices produce models that are accurate
but catastrophically overconfident (we observed a 0.64-to-0.35 log loss
improvement from this change alone). Second, post-hoc parameters inherit
the regime of the slice they are fit on: the same blend weights that
help in the temporal regime hurt in the interleaved one. There is no
regime-free calibration.

\section{Results I: The Regime Audit}
\label{sec:audit}

Table~\ref{tab:regime} is the paper's central exhibit: one set of
models, three forecasting regimes, log loss and accuracy side by side.
Table~\ref{tab:regimeA} reports the completion regime separately, since
its models observe same-rollcall votes.

\begin{table}[t]
\centering
\caption{The evaluation-regime audit. Test log loss (left panel) and
accuracy (right panel) under random rollcall holdout, within-congress
temporal forecast, and whole-congress-out transfer. Same-day paired
fits; each test set observed once for the final comparison.}
\label{tab:regime}
\input{tables/regime_audit.tex}
\end{table}

\begin{table}[t]
\centering
\caption{The completion regime (random held-out votes, other votes on
the same rollcall observable): the psychometric setting in which
ideal-point models are traditionally evaluated.}
\label{tab:regimeA}
\input{tables/regimeA.tex}
\end{table}

Four patterns organize everything else. First, completion is easy and
essentially solved: an eight-dimensional ideal-point model reaches
\regimeAbestLL{} log loss (\regimeAbestAcc\% accuracy), and even one
dimension nearly matches DW-NOMINATE's decades of refinement. Reading
completion-regime numbers as ``prediction'' wildly overstates what any
model can do about the future.

Second, among true forecasting regimes, the literature's random-rollcall
choice was more defensible than we expected. Raw numbers differ between
random-rollcall and temporal splits, but so do the test sets'
compositions (end-of-congress rollcalls are more lopsided); measured as
lift over the same-regime null, the two regimes agree almost exactly.
The audit's sharp cliffs are elsewhere: between completion and any
rollcall-holdout regime, and between within-congress and cross-congress.

Third, under proper scoring the classical text models were never
competitive. The raw Gerrish--Blei-style model posts a forecast log loss
of \gbRawForecastLL---worse than the constant baseline---while its
accuracy looks acceptable; its discrimination (AUC \gbForecastAUC{}
versus the champion's \champForecastAUC) and its calibration are both at
fault, and repairing calibration still leaves it behind a count table.
The Kraft-style bilinear model, given every modern courtesy, reaches
\kraftForecastLL{} and also trails the member-by-question baseline. The
2011--2016 program, in short, demonstrated that text \emph{correlates}
with voting; it did not demonstrate forecasting value over member
history, because it never raced against member history under a proper
score.

Fourth, the modern stack clears all of it by a wide margin, and the
final blend adds a further \champGain{} log loss over the best single
tower, with its validation-predicted gain reproducing almost exactly on
the once-observed test set.

\section{Results II: What Moves the Needle, and What Does Not}

Table~\ref{tab:ablation} traces the path from a metadata-only tower to
the champion. The large steps are rollcall-level text construction
(giving content to nominations and amendments rather than reusing parent
bill text), the member-by-question offset, the deep parameter head, and
temporal calibration. The small or negative steps are equally
instructive, and we report them with mechanisms rather than as bare
nulls; the full pre-registered ledger is in the appendix.

\begin{table}[t]
\centering
\caption{Ablation path in the temporal forecast regime (test set).
Each row modifies the configuration above it; the final rows blend
towers.}
\label{tab:ablation}
\input{tables/ablation.tex}
\end{table}

\begin{table}[t]
\centering
\caption{The literature horse race in detail: forecast-regime log loss,
AUC, and contested-vote accuracy, with random-rollcall figures for
comparison.}
\label{tab:lit}
\input{tables/litrace.tex}
\end{table}

\textbf{Bigger encoders do not help.} Replacing a 384-dimensional
MiniLM encoder with a 1,024-dimensional long-context model (a
sixteen-fold larger text budget) left the forecast essentially
unchanged and made long-bill votes slightly \emph{worse}---rejecting the
natural hypothesis that summary truncation was the binding constraint.
Omnibus coalitions are assembled around deals whose terms are not in the
text. The larger encoder's tower does, however, earn a place in the
final blend: its errors are decorrelated from the MiniLM tower's, which
is worth more than its solo performance.

\textbf{Amendment text does not help, and we can say why.}
Amendment votes carry roughly thirty percent of residual loss, and
two-thirds of amendment rollcalls have no amendment-specific text---so
enriching them looked like the obvious next win. It is not. We parsed
purpose language for every amendment in the BILLSTATUS corpus, joined it
to rollcalls through recorded-vote references (correcting a
session-numbering trap that silently mismatches forty percent of joins),
and rebuilt the text pipeline three ways, from aggressive to minimal.
Every variant made the model \emph{worse}, by amounts insensitive to
whether the joined text was correct; a placebo rebuild that changed the
file without changing any content reproduced the champion exactly,
exonerating the pipeline mechanics. The mechanism: when a bill's
amendments share the parent bill's text, the model assigns the whole
series nearly identical parameters---an implicit pooled estimate of a
highly correlated block of votes. Distinct purpose text replaces that
pooling with per-amendment parameters driven by content, and content
predicts amendment coalitions worse than the bill's identity does. The
error decomposition (Table~\ref{tab:decomp}) locates the loss; the
experiments show it is not a text problem.

\begin{table}[t]
\centering
\caption{Where the loss lives: test log loss by question type, prior
champion versus final blend, temporal forecast regime.}
\label{tab:decomp}
\input{tables/error_decomp.tex}
\end{table}

\textbf{Member-history refinements do not help.} Recency-weighted member
rates (halving the effective sample behind each cell) and within-bill
context features (a member's vote on the same bill's previous rollcall)
both failed pre-registered validation tests---the latter despite passing
its synthetic checks at the information-theoretic optimum, because the
real-data signal available under a strict no-leakage harness is small
and its exploitation destabilizes the rest of the model.

\section{Results III: The Transfer Wall}
\label{sec:transfer}

The congress-out column of Table~\ref{tab:regime} is the paper's
sobering exhibit. Train through 2021, predict 2023--24---across the
House majority flip---and the champion's log loss more than doubles
while its accuracy falls below every within-congress baseline. Three
further facts sharpen the diagnosis.

First, the honest floor for this regime is a count table keyed on
majority status, chamber, and question type, all knowable before any
vote: it scores \majorityCongressoutLL{}---\emph{better calibrated than
the champion's} \champCongressoutLL. Second, member history is not
merely unhelpful across the flip; it is poisonous. A pooled
member-by-question table anti-transfers (log loss 0.740), because
procedural voting is majority-role behavior: the member who reliably
voted yea on previous questions as a majority member votes nay on them
in the minority, and his history argues confidently for the wrong side.
We verify the mechanism in a synthetic two-world experiment: where
behavior is majority-driven, member history anti-transfers exactly as
observed; where behavior is a stable personal trait, it transfers
nearly perfectly. Real congressional behavior is evidently a blend, with
the majority-role component dominating procedural votes. Third, simply
handing the model majority-status features does not fix it---within a
congress the member offsets are strictly more informative, so training
assigns the majority features no weight, and at transfer time they
cannot substitute for the collapsed offsets. A genuine solution likely
requires majority-conditioned training, which we leave as an open
problem posed to the benchmark. Meanwhile, a logit on frozen DW-NOMINATE
scores---positions estimated from full careers spanning both majority
roles---very nearly matches the champion's transfer accuracy
(\nomctxCongressoutAcc\% versus \champCongressoutAcc\%), a quiet
vindication of career-spanning measurement precisely where short-window
learning fails.

\section{Results IV: The Prospective Test}

Development test sets, however carefully rationed, are observed by the
researchers iterating on them. Our headline forecasting claim therefore
rests elsewhere. On June 12, 2026, we froze the then-champion single
tower---weights pickled, SHA-256 recorded---fit on all votes through
June 9. Every vote cast afterward constitutes an evaluation set that did
not exist when the model was finalized. Through June 30, that is
\prospectiveRC{} rollcalls and \prospectiveN{} individual votes: log
loss \prospectiveVoneLL, accuracy \prospectiveVoneAcc\%---statistically
indistinguishable from the model's development-test performance, which
is the entire point: the development numbers in this paper are not
inflated by adaptive overfitting. The blend champion, frozen July 3, has
begun its own clock (identical log loss, one point higher accuracy, on
a window too small to separate them); both artifacts remain pinned and
will accumulate evidence as Congress votes.
Table~\ref{tab:prospective} reports the ledger.

\begin{table}[t]
\centering
\caption{Frozen prospective artifacts, scored only on votes occurring
after each model's data snapshot. Cluster-bootstrap intervals in the
appendix.}
\label{tab:prospective}
\input{tables/prospective.tex}
\end{table}

\section{Implications}

For measurement, the results cut two ways. The completion regime's
near-ceiling performance and the transfer regime's collapse together
imply that vote-based scores are exquisitely good summaries of behavior
\emph{within} an institutional configuration and should be handled with
care \emph{across} them; the surprising cross-congress durability of
career-spanning NOMINATE scores, relative to far richer short-window
models, is a genuine argument for the traditional approach's design.
For the text-as-data program, the lesson is that representation was
never the bottleneck in this domain: what modern models add is mostly
better use of behavioral history and better-calibrated uncertainty, and
the marginal value of reading legislative language, while real, is
concentrated in bills whose content actually organizes the coalition.
For forecasting practice, temporal calibration and regime-matched
validation slices matter more than architecture, and probabilities
reported without a proper-scoring audit should be treated as
decorative.

A companion paper uses the champion's internals for substantive
measurement---party-loyalty residuals, issue-specific positions, and
out-of-character votes. The benchmark, leaderboard, frozen artifacts,
and complete experiment ledger (including every negative result
reported above, pre-registered before its run) are public; the transfer
problem in Section~\ref{sec:transfer} is offered to the field as the
benchmark's standing challenge.

\bibliographystyle{apsr}
\bibliography{references}

\end{document}
